---
title: "Scientific background for fieldClim heat-flux methods"
output: rmarkdown::html_vignette
bibliography: fieldclim-methods.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Scientific background for fieldClim heat-flux methods}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Purpose of this vignette

This vignette explains the scientific background of the heat-flux methods implemented in `fieldClim`. It is not a numerical benchmark, and it does not claim that one method is the universal reference for all surfaces, time scales and stability regimes. Its purpose is to make the method logic explicit: which part of the surface energy balance is used, which station variables are required, which assumptions enter the calculation, and how the resulting fluxes should be interpreted. The methodological frame is the same one used in micrometeorology and environmental biophysics: turbulent fluxes can be measured directly with covariance systems, estimated with energy-balance and aerodynamic methods, or diagnosed from gradients and stability assumptions [@oke1987; @monteith2013; @foken2008book; @campbell1998].

`fieldClim` is designed for weather stations and microclimate stations that measure the near-surface drivers of the energy balance but do not include full eddy-covariance instrumentation. Typical inputs are net radiation or radiation components, soil heat flux or soil-temperature information, air temperature, humidity and wind at one or more heights. Such stations cannot produce direct covariance fluxes from high-frequency vertical wind and scalar fluctuations. They can, however, support established approximation families that diagnose or estimate sensible and latent heat fluxes from mean meteorological quantities [@allen1998; @brutsaert1982; @prueger2005; @campbell1998].

The central point is therefore not to defend one particular method. Priestley-Taylor, Penman-type, Bowen-ratio, bulk-resistance/residual and Monin-Obukhov/profile approaches belong to different micrometeorological method families, and they answer different operational questions [@penman1948; @priestley1972; @allen1998; @prueger2005; @stull1988]. Some paths close the available energy by construction, one path provides only latent heat flux, and profile methods are diagnostic rather than energy-closing in this package. This difference is the reason why `fieldClim` implements several paths side by side.

# Measuring the surface energy balance: from eddy covariance to station-based methods

Eddy covariance estimates turbulent fluxes from the covariance between fluctuations in vertical wind velocity and fluctuations in scalar quantities such as temperature, water vapour or trace-gas concentration [@aubinet2012; @foken2012ec; @lee2004; @kaimal1994]. In a full EC system, a three-dimensional sonic anemometer provides the turbulent wind components and often sonic temperature for sensible heat flux, while a fast-response gas analyser provides water-vapour or CO2 fluctuations for latent heat and trace-gas fluxes [@aubinet2012; @lee2004]. This is a fundamentally different data type from ordinary weather-station data, which usually provide averaged radiation, temperature, humidity, wind and soil measurements rather than high-frequency turbulent covariances.

This distinction matters for `fieldClim`. The package does not process high-frequency EC data. It does not rotate wind vectors, despike raw turbulence records, correct time lags, estimate spectral losses, calculate coordinate transformations or perform EC quality control. Those operations belong to the EC processing chain described in EC handbooks and method chapters, not to the mean-station data model used here [@aubinet2012; @foken2012ec; @lee2004].

EC nevertheless remains essential background because many simpler energy-balance, gradient, aerodynamic and evaporation-model approaches are evaluated against EC or sonic-anemometer based flux observations. At the same time, EC is not an error-free truth source. The energy-balance closure problem is well documented, and closure depends on footprint, heterogeneity, averaging time, storage terms, advection, instrumentation and data filtering [@foken2008closure; @aubinet2012]. For `fieldClim`, EC comparison studies therefore provide empirical context, not an automatic target that every station-based method must reproduce exactly.

## What intercomparison studies actually show

The intercomparison literature does not support a simple hierarchy in which EC is always truth and all station methods are merely crude substitutes. It shows a more useful pattern: direct turbulent-flux systems, Bowen-ratio systems, residual-energy approaches, Penman-type models, Priestley-Taylor estimates, aerodynamic/profile methods and bulk-transfer models are repeatedly compared because each method becomes useful under different instrumentation, surface and stability conditions [@billesbach2024; @shi2008; @sumner2005; @tomlinson1996].

The studies below are included because they connect directly to the method families used in `fieldClim`. The table is not a complete literature review. It is a map of the evidence base that explains why the package contains several method families rather than one single flux estimator.

| Evidence block | Key references | What it supports for `fieldClim` |
|---|---|---|
| EC, energy-balance, residual and Bowen-type comparison over grassland | @billesbach2024 | Direct comparison of EC, energy-balance/Bowen-ratio, residual-energy and modified Bowen-ratio methods for sensible and latent heat flux over grassland. |
| EC, Bowen-ratio energy balance and Penman-Monteith comparison | @shi2008 | Shows that EC, BREB and Penman-Monteith are evaluated as different operational pathways, not as identical measurements. |
| Bowen-ratio, eddy-correlation and lysimeter comparison | @tomlinson1996 | Links Bowen-ratio and eddy-correlation estimates with an independent evapotranspiration reference, and documents practical field-method differences. |
| Penman-Monteith and aerodynamic resistance from meteorological data | @allen1998 | Provides the standard meteorological basis for Penman-Monteith, surface resistance, aerodynamic resistance and wind-dependent reference formulations. |
| PT and PM evapotranspiration comparison | @sumner2005; @priestley1972; @penman1948 | Places Priestley-Taylor and Penman-type estimates in the practical evapotranspiration-method family. |
| Aerodynamic and flux-gradient methods | @prueger2005 | Supports profile-gradient and aerodynamic transfer methods as an established micrometeorological method family for turbulent flux estimation. |
| Bulk-transfer estimates against sonic/scintillometer-type observations | @kim2019 | Supports empirical comparison of bulk-transfer sensible heat estimates against direct or near-direct turbulent-flux observations. |
| Bulk-transfer models against EC over several seasons | @alakonni2022 | Shows that bulk-transfer models are evaluated against EC over multi-season field data, although surface type and transfer coefficients differ from the Caldern meadow case. |
| EC closure limitation | @foken2008closure | Prevents treating EC as an error-free truth source in method interpretation. |
| Boundary-layer and surface-exchange theory | @stull1988; @garratt1992; @kaimal1994; @arya2001 | Provides the general theory context for gradients, turbulence, stability and surface-layer exchange. |

The conclusion is not that one method is universally correct. The conclusion is that the implemented approaches belong to evaluated micrometeorological method families. Their differences are not a weakness of the package; they are the reason the methods must be shown separately. Each method exposes a different compromise between data demand, physical detail, numerical robustness and closure behaviour.

## Method families and their assumptions

The method families in `fieldClim` differ mainly in how they use the surface energy balance. Priestley-Taylor, Bowen and Bulk-Residual explicitly partition or close the available energy. Penman provides only a latent heat estimate. Monin-Obukhov/profile methods use profile and stability information and are therefore diagnostic rather than energy-closing in this package. The Bulk-Residual path now also has an optional Richardson guard; this guard screens the neutral sensible-heat estimate but does not turn the method into a MO correction.

| Method family | `fieldClim` functions | How station data enter | Closure behaviour | Main sensitivity | Literature anchors |
|---|---|---|---|---|---|
| Priestley-Taylor | `latent_priestley_taylor()`, `sensible_priestley_taylor()` | Available energy and empirical PT coefficient. | \(H + LE\) closes \(R_n - G\) by construction. | PT coefficient and available-energy quality. | @priestley1972; @sumner2005 |
| Penman-type | `latent_penman()` | Available energy plus aerodynamic vapour-pressure term. | LE only; no paired H is returned. | Humidity, wind, resistance assumptions and pressure-unit consistency. | @penman1948; @allen1998; @brutsaert1982 |
| Bowen ratio | `sensible_bowen()`, `latent_bowen()` | Potential-temperature and absolute-humidity gradients partition available energy through the implemented `fieldClim` beta path. | Closes \(R_n - G\) only for finite uncapped denominators; capped or non-finite beta cases are guarded diagnostics. | Small humidity gradients, sign changes, non-finite beta, near-zero denominator and source-form ambiguity of `gamma_code`. | @tomlinson1996; @shi2008; @billesbach2024 |
| Bulk-Residual | `sensible_bulk()`, `latent_bulk_residual()`, `turb_flux_bulk_residual()` | Temperature gradient and wind-dependent resistance estimate H; optional Richardson guard screens very stable or weak-shear cases; residual closes LE. | Closes \(R_n - G\) by construction when H is valid; guarded H values propagate to NA residual LE. | Wind representation, temperature gradient, Richardson screening, weak-shear cases and residual error propagation. | @prueger2005; @kim2019; @alakonni2022; @allen1998 |
| Monin-Obukhov / profile | `sensible_monin()`, `latent_monin()`, `turb_flux_stability()` | Profile gradients, wind information and stability logic diagnose H and LE. | Diagnostic only; not forced to close \(R_n - G\). | Low wind, zero gradients, weak wind shear, stability classification and invalid height/profile states; guarded cases return diagnostic `NA`. | @stull1988; @garratt1992; @foken2008book; @prueger2005 |
| Radiation and soil helpers | `rad_*()`, `sol_*()`, `trans_*()`, `soil_*()` | Provide radiation, soil and auxiliary variables used by flux methods. | Input and consistency layer, not a turbulent-flux closure method. | Sensor consistency, surface type, time base, terrain and soil assumptions. | @oke1987; @campbell1998; @monteith2013; @bendix2004 |

This table should be read from left to right. The package functions are not alternative names for the same measurement. They are different ways of turning available station information into heat-flux estimates or diagnostics. Their outputs therefore have different closure behaviour and different failure modes.

## Why fieldClim implements these methods and not EC processing

`fieldClim` starts from the instrumentation reality of many field stations: radiation and soil variables are available, temperature and humidity profiles may be available, wind speed may be available at one or two heights, but high-frequency turbulence data usually are not. Under these conditions, EC processing would be the wrong package scope. It would require a different data model, a different quality-control chain and different raw variables [@aubinet2012; @foken2012ec; @lee2004].

The implemented methods are therefore not second-best versions of an EC workflow. They are appropriate method families for mean station data. PT uses available energy with a compact coefficient; Penman uses available energy and aerodynamic vapour-pressure forcing; Bowen uses gradients to partition available energy; Bulk-Residual combines a wind-dependent sensible heat estimate, optionally screened by a Richardson guard, with residual latent heat; and MO/profile methods diagnose turbulent exchange from vertical gradients and stability information [@priestley1972; @penman1948; @allen1998; @prueger2005; @stull1988]. These approaches are comparable in the literature because they are used when the available data, surfaces and time scales differ [@billesbach2024; @shi2008; @sumner2005; @alakonni2022].

# Surface energy balance and package sign convention

The common reference point is the surface energy balance. Microclimatological and micrometeorological texts use partly different symbols for the same physical quantities, but the underlying balance is shared: net radiation is partitioned into ground heat flux, sensible heat flux, latent heat flux and, where relevant, storage or advection terms [@oke1987; @monteith2013; @foken2008book; @bendix2004]. `fieldClim` uses technical field and output names such as `rad_bal`, `soil_flux`, `sensible_*` and `latent_*`. These are implementation names, not a separate theoretical notation.

```{r sign-convention-table, echo = FALSE}
sign_convention <- data.frame(
  Quantity = c(
    "Net radiation",
    "Soil heat flux",
    "Sensible heat flux",
    "Latent heat flux",
    "Available turbulent energy"
  ),
  `Common notation` = c(
    "Q* or Rn",
    "B or G",
    "H or L",
    "LE or V",
    "Q* - B or Rn - G"
  ),
  `fieldClim field / output` = c(
    "rad_bal",
    "soil_flux",
    "sensible_*",
    "latent_*",
    "rad_bal - soil_flux"
  ),
  `Positive direction / meaning in fieldClim` = c(
    "net radiative energy input at the surface",
    "heat flux into the soil",
    "heat flux away from the surface",
    "heat flux away from the surface",
    "energy available for sensible and latent heat fluxes"
  ),
  check.names = FALSE
)

knitr::kable(sign_convention)
```

With storage omitted, the surface energy balance is written here as:

$$
R_n = G + H + LE
$$

Therefore the energy available for turbulent heat fluxes is:

$$
R_n - G = H + LE
$$

In the notation used in the accompanying theory material, the same balance is:

$$
Q^* - B = L + V
$$

A residual latent heat flux follows directly from the same balance:

$$
LE = R_n - G - H
$$

This convention is used throughout the package: `soil_flux` is consumed as a positive heat flux into the soil, while positive sensible and latent heat fluxes are interpreted as fluxes away from the surface. This convention is also why the residual methods are algebraically transparent: if \(R_n\), \(G\) and \(H\) are specified, \(LE\) follows from the balance.

# Method-specific background

## Priestley-Taylor

The Priestley-Taylor method reduces evaporation and latent heat estimation to an energy-parameter problem. It starts from available energy and scales the equilibrium evaporation term with an empirical coefficient [@priestley1972]. This makes it attractive when radiation and soil heat flux are available but profile-gradient information is incomplete or noisy. The same method family is commonly compared with Penman-Monteith and other evapotranspiration estimates in practical ET applications [@sumner2005].

The implemented PT path has the structure:

$$
LE_{PT} =
\alpha
\frac{\Delta}{\Delta + \gamma}
(R_n - G)
$$

The corresponding sensible heat flux is calculated as the remaining available energy:

$$
H_{PT} = (R_n - G) - LE_{PT}
$$

Therefore the closure invariant is exact:

$$
H_{PT} + LE_{PT} = R_n - G
$$

The practical strength of PT is robustness. It needs fewer profile assumptions than Bowen or MO and avoids explicit wind-resistance modelling. The price is that surface-atmosphere coupling is condensed into the coefficient and helper terms. PT is therefore useful as a stable baseline for station-based energy partitioning, not as a direct measurement of turbulence [@priestley1972; @sumner2005].

## Penman-type latent heat path

The Penman family is a combination family: it links the energy available for evaporation with an aerodynamic term driven by atmospheric demand [@penman1948]. FAO-56 later standardized the Penman-Monteith framework for crop evapotranspiration and explicitly distinguishes the energy term, surface resistance and aerodynamic resistance [@allen1998]. In `fieldClim`, `latent_penman()` is intentionally LE-only. It does not return a paired sensible heat flux.

The package keeps the energy term as:

$$
R_n - G
$$

The implemented Penman path uses saturation vapour pressure and actual vapour pressure helpers that return hPa, converts them internally to kPa for the aerodynamic vapour-pressure deficit, and keeps the saturation-curve slope and psychrometric constant on the same pressure scale:

$$
VPD_{kPa} = \frac{e_s - e_a}{10}
$$

This unit consistency matters because mixing hPa vapour-pressure deficit with kPa-scale helper terms inflates the aerodynamic contribution. After the unit correction, the method remains a modelled latent heat estimate. It should not be compared with PT, Bowen or Bulk-Residual as if it also delivered a corresponding sensible heat flux. The general combination-method logic belongs to the Penman/Penman-Monteith family, but the package path is a simplified latent-heat implementation from standard station variables [@penman1948; @allen1998; @brutsaert1982].

The relevant package distinction is:

$$
\text{Penman path} = LE \text{ only}
$$

There is no package-defined Penman sensible heat flux.

## Bowen-ratio

The Bowen-ratio method partitions available energy using the ratio of sensible to latent heat flux:

$$
\beta = \frac{H}{LE}
$$

If the ratio is known, the energy balance gives:

$$
H =
\frac{\beta}{1 + \beta}
(R_n - G)
$$

and:

$$
LE =
\frac{1}{1 + \beta}
(R_n - G)
$$

For finite uncapped denominators, the closure invariant is exact:

$$
H + LE = R_n - G
$$

The Bowen-ratio energy-balance family is widely used because it connects measured vertical gradients to energy partitioning and can be compared directly with EC or lysimeter-based evapotranspiration estimates [@tomlinson1996; @shi2008; @billesbach2024]. In `fieldClim`, the implemented Bowen ratio is based on a potential-temperature gradient and an absolute-humidity gradient:

$$
\beta =
\gamma_{code}
\frac{\Delta \theta / \Delta z}{\Delta AH / \Delta z}
$$

with:

$$
\gamma_{code} =
0.00066
(1 + 0.000946 t_1)
$$

This matters because the implementation is more specific than a symbol-only textbook expression. The package converts `t1` and `t2` to potential temperature, converts `hum1` and `hum2` from relative humidity to absolute humidity, and then forms the gradient ratio. The coefficient \(\gamma_{code}\) is therefore treated as the empirical coefficient of the current `fieldClim` implementation. Current validation did not prove source-equivalence between this form and alternative Bowen-ratio formulations based on vapour-pressure gradients, specific-humidity gradients or heat-capacity / latent-heat scaling. The implemented beta path is therefore tested and documented as package behaviour, while the exact source-form decision remains open.

The method is powerful when gradients are well resolved. The same dependence makes it fragile. Small humidity gradients, sign changes and near-zero values of `1 + beta` can dominate the result. The `cap` parameter is therefore a numerical safeguard for near-zero denominators. For finite uncapped cases, `sensible_bowen()` and `latent_bowen()` use the same beta path and close \(R_n - G\). If the denominator is capped, or if beta or the denominator becomes non-finite, the output is treated as guarded diagnostic behaviour rather than as an exact physical closure. Non-finite Bowen ratios or denominators are returned as `NA` for the affected time steps.

## Bulk-Residual

The Bulk-Residual path combines an aerodynamic-resistance estimate of sensible heat flux with a residual latent heat flux. It is not an EC measurement and it is not a full MO surface-layer model. It is a station-data implementation of a familiar aerodynamic-transfer idea: wind-driven exchange controls the resistance between a temperature difference and a sensible heat flux [@prueger2005; @allen1998; @campbell1998].

The sensible heat calculation follows the general aerodynamic-resistance form:

$$
H_{bulk} =
\rho c_p
\frac{\Delta T}{r_a}
$$

In the neutral default implementation:

$$
\Delta T = T_1 - T_2
$$

and:

$$
r_a =
\frac{\ln(z_2 / z_1)}
{k \bar{u}}
$$

Here \(T_1\) and \(T_2\) are lower and upper air temperatures, \(z_1\) and \(z_2\) are the corresponding measurement heights, \(k\) is the von Karman constant and \(\bar{u}\) is the observed wind-speed scale used by the package. If two wind heights are available, \(\bar{u}\) is the arithmetic mean of `v1` and `v2`; otherwise `v1` is used.

The wind term belongs to the aerodynamic-resistance logic used in micrometeorology, evapotranspiration modelling and flux-gradient methods [@allen1998; @prueger2005; @monteith2013]. It should not be read as a full friction-velocity-based surface-layer formulation. More complete formulations may use friction velocity, roughness length, displacement height and stability corrections [@stull1988; @garratt1992; @foken2008book]. The default `fieldClim` implementation is deliberately simpler because standard station datasets often provide wind speed but not directly measured friction velocity.

The optional Richardson guard does not replace this neutral bulk calculation. It screens the neutral estimate when two-height wind and temperature information are available. With `stability_method = "ri_guard"`, the package estimates a gradient Richardson number:

$$
Ri_g =
\frac{g}{\bar{\theta}}
\frac{\Delta \theta / \Delta z}
{(\Delta u / \Delta z)^2}
$$

where \(g\) is gravitational acceleration, \(\bar{\theta}\) is mean potential temperature if elevation is available, \(\Delta \theta / \Delta z\) is the vertical potential-temperature gradient, and \(\Delta u / \Delta z\) is the vertical wind-speed shear between the two measurement heights. If elevation is not available, the implementation uses Kelvin air temperature as a near-surface approximation for the stability screen.

This Richardson number is used as a guard, not as a flux correction. Unstable, neutral and moderately stable cases keep the neutral bulk estimate. Very stable cases, invalid Richardson numbers and weak-shear cases are returned as `NA` for the affected time steps. The guard therefore prevents a formally computed neutral bulk flux from being treated as robust when the two-height profile indicates that the neutral approximation is not defensible.

The implementation uses wind-speed shear rather than full vector shear because ordinary station data usually provide wind speed but not horizontal wind components. This makes the guard a speed-shear Richardson approximation. It is useful for screening problematic time steps, but it is not a full MO correction and it does not apply stability functions to rescale the flux.

The residual latent heat flux is:

$$
LE_{res} =
R_n - G - H_{bulk}
$$

Therefore, when \(H_{bulk}\) is valid, the Bulk-Residual workflow closes available energy by construction:

$$
H_{bulk} + LE_{res} = R_n - G
$$

If the Richardson guard sets \(H_{bulk}\) to `NA`, the residual \(LE_{res}\) is also `NA` for that time step. This is intended. It prevents residual closure from hiding an invalid or non-robust sensible heat estimate.

This closure is algebraic. It does not prove that `H_bulk` is a perfect physical estimate. It means that the residual latent heat flux inherits all errors in \(R_n\), \(G\) and the bulk sensible heat estimate. This is not a special weakness of the Bulk-Residual path. It is the normal consequence of residual energy-balance reasoning.

Empirically, the broader bulk-transfer and aerodynamic-transfer family has been compared with direct or near-direct turbulent-flux observations. Bulk-transfer sensible heat estimates have been validated against three-dimensional ultrasonic anemometer or scintillometer-based observations in UAV-based work [@kim2019], and bulk-transfer models have been evaluated against EC turbulent heat fluxes over seasonal lake ice [@alakonni2022]. These studies do not validate every possible station implementation, but they show that bulk/aerodynamic transfer is an evaluated method family rather than a package invention.

### Wind speed, friction velocity and two-height station data

The use of wind is not an add-on. Stronger wind-driven turbulent exchange lowers aerodynamic resistance, so the same temperature difference produces a larger sensible heat flux when mixing is stronger. This wind dependence is explicit in aerodynamic-resistance formulations used in Penman-Monteith and related station-data methods [@allen1998; @monteith2013].

If only one wind speed is available, a friction-velocity-based method cannot be used without additional roughness assumptions. If two wind heights are available, friction velocity can be estimated under a neutral log-profile assumption:

$$
u_* =
k
\frac{u_2 - u_1}
{\ln(z_2 / z_1)}
$$

and then a neutral heat-transfer resistance can be written as:

$$
r_{ah} =
\frac{\ln(z_2 / z_1)}
{k u_*}
$$

This is a possible alternative formulation, but it is not automatically more robust for ordinary station data. If the wind-speed difference is small, noisy or changes sign, the estimated friction velocity becomes unstable. The current Bulk-Residual implementation therefore keeps the observed-wind neutral resistance as its default and uses the Richardson number only as an optional stability guard.

The important distinction is this: `sensible_bulk()` is not a full friction-velocity or MO method. Its default path is a neutral bulk-resistance estimate from observed wind speed. Its optional Richardson guard is a diagnostic screen for stability and weak-shear problems. A future friction-velocity-based mode would be a separate method option and should not silently replace the current default.

## Monin-Obukhov and profile methods

Monin-Obukhov similarity theory and related profile-gradient methods represent turbulent transfer through vertical gradients, roughness, stability and surface-layer scaling [@stull1988; @garratt1992; @foken2008book]. They are methodologically important because they connect observed profiles with turbulence theory. In `fieldClim`, however, the MO path is diagnostic-only.

The package uses profile information to derive diagnostic sensible and latent heat-flux estimates from vertical temperature, humidity and wind information. The sensible heat profile term follows the documented vertical-gradient structure and uses the height difference \(z_2 - z_1\). This keeps the implemented profile-gradient term aligned with the Richardson-gradient logic used elsewhere in the package.

The diagnostic rule remains:

$$
H_{MO} + LE_{MO}
$$

is not required to equal:

$$
R_n - G
$$

This is not a defect. It is a consequence of method class. A profile method estimates fluxes from profile and stability assumptions, while PT, Bowen and Bulk-Residual are explicit energy-partition or residual workflows. Forcing MO output to close \(R_n - G\) would hide the diagnostic information that profile methods are meant to expose.

The practical issue is numerical robustness. Zero gradients, very small wind shear, invalid height relationships and unstable stability functions can produce misleading or undefined outputs. `fieldClim` therefore treats these cases explicitly. A zero potential-temperature gradient returns zero sensible heat flux. A zero moisture gradient returns zero latent heat flux. Invalid heights, invalid wind and invalid numerical profile states return `NA` with a warning. Richardson-number diagnostics return `NA` for invalid heights, invalid wind, weak or zero shear, and non-finite values.

This guard behaviour does not make MO an energy-balance closure method. It only prevents invalid profile states from being interpreted as ordinary heat-flux events. MO outputs should therefore be inspected together with stability classification and diagnostic warnings. They should not be silently normalized to available turbulent energy.

# Practical interpretation in fieldClim

The practical interpretation follows from the closure behaviour. If the goal is a stable first estimate of the partitioning of available energy, the PT path is the least demanding because it avoids explicit profile-gradient and wind-resistance assumptions. Its stability is bought by using an empirical coefficient. It is therefore useful as a baseline, not as a direct turbulent-flux measurement [@priestley1972; @sumner2005].

The Penman path should be interpreted separately because it is LE-only. It combines an energy term with an aerodynamic vapour-pressure term and is useful when a latent heat estimate is needed from standard meteorological variables. It should not be compared with PT, Bowen or Bulk-Residual as if it also provided a corresponding sensible heat flux [@penman1948; @allen1998].

The Bowen path is a gradient-partitioning method. Its strength is that it uses measured vertical gradients to divide available energy into sensible and latent parts. Its weakness is the same property: small humidity gradients, sign changes and near-singular values of `1 + beta` can dominate the result. In `fieldClim`, the beta path is locked and tested as the current implementation, but the exact source-form equivalence of the empirical \(\gamma_{code}\) coefficient remains open. Exact closure is therefore a property of finite uncapped cases, not of capped, non-finite or otherwise guarded Bowen outputs [@tomlinson1996; @shi2008; @billesbach2024].

The Bulk-Residual path uses measured temperature gradients and wind speed to estimate sensible heat flux, optionally screens this estimate with the Richardson guard, and then assigns the remaining available energy to latent heat. This makes the residual closure transparent, but the closure is only meaningful for time steps where the sensible heat estimate is valid. If the guard marks a very stable, weak-shear or otherwise invalid profile as non-robust, `H_bulk` and the residual `LE_res` become `NA` for that time step. The method should therefore be interpreted as a station-data aerodynamic/residual pathway with an optional stability screen, not as a full MO correction [@prueger2005; @kim2019; @alakonni2022; @stull1988; @garratt1992].

The MO/profile path has the strongest link to profile and stability reasoning, but it is also the most numerically sensitive. The current implementation guards zero-gradient, low-wind, invalid-height and non-finite Richardson cases, but it remains diagnostic-only. Its value is to show how profile gradients and stability assumptions behave in the available data, not to force closure with \(R_n - G\). Guarded `NA` values should therefore be read as diagnostic information rather than ordinary heat-flux events [@stull1988; @garratt1992; @foken2008book; @arya2001].

The central conclusion is therefore not that one method is universally correct. The implemented methods belong to evaluated micrometeorological method families, but they answer different operational questions. `fieldClim` makes these differences explicit so that station-based heat-flux estimates can be compared, diagnosed and interpreted without pretending that they are interchangeable measurements.

# References
