---
title: "fieldClim: Check radiation and soil heat flux"
subtitle: "Geländeklimatologie SoSe 2023, Philipps-Universität Marburg"
author: "You-De Chen"
date: "`r format(Sys.time(), tz = 'UTC', usetz = TRUE)`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{fieldClim: Check radiation and soil heat flux}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
caldern_path <- system.file("extdata", "caldern_wiese_2017-06-30.csv",
                            package = "fieldClim")
```

```{r}
# load library
library(fieldClim)
```

# check .default methods
```{r}
datetime <- as.POSIXlt("2018-02-19 13:15:00", tz = "CET")
lon = 8.683300
lat = 50.840503
elev = 270
temp = 20
rh = 50
slope = 30
exposition = 20
valley = FALSE
surface_type = "field"
surface_temp = 20
texture = "sand"
moisture = 0.2
soil_temp1 = 20
soil_temp2 = 30
soil_depth1 = 1
soil_depth2 = 0

# structure
#*1 means there are optional arguments
#*1.0 means there are optional arguments originated from this function

## soil
soil_heat_flux(texture, moisture, soil_temp1, soil_temp2, soil_depth1, soil_depth2)
  soil_thermal_cond(texture, moisture)

## rad
rad_bal(datetime, lon, lat, elev, temp, rh, slope, exposition, valley, surface_type, surface_temp)*1
  rad_sw_bal(datetime, lon, lat, elev, temp, slope, exposition, valley, surface_type)*1
    rad_sw_out(datetime, lon, lat, elev, temp, slope, exposition, surface_type)*1
      rad_sw_in(datetime, lon, lat, elev, temp, slope, exposition)*1
        rad_sw_toa(datetime, lon, lat)*1.0
          sol_eccentricity(datetime)
            sol_day_angle(datetime)
              sol_julian_day(datetime)
          sol_elevation(datetime, lon, lat)
            sol_declination(datetime)
              sol_ecliptic_length(datetime)
                sol_medium_anomaly(datetime)
            sol_hour_angle(datetime, lon)
              sol_medium_suntime(datetime, lon)
              sol_time_formula(datetime, lon)
        trans_gas(datetime, lon, lat, elev, temp)*1
          trans_air_mass_abs(datetime, lon, lat, elev, temp)*1
            trans_air_mass_rel(datetime, lon, lat)
            pres_p(elev, temp)*1
        trans_ozone(datetime, lon, lat)*1.0
        trans_rayleigh(datetime, lon, lat, elev, temp)*1
        trans_vapor(datetime, lon, lat, elev, temp)*1
          hum_precipitable_water(datetime, lat, elev, temp)*1
        trans_aerosol(datetime, lon, lat, elev, temp)*1.0
        terr_terrain_angle(datetime, lon, lat, slope, exposition)
          sol_azimuth(datetime, lon, lat)
    rad_diffuse_out(datetime, lon, lat, elev, temp, slope, exposition, valley, surface_type)*1
      rad_diffuse_in(datetime, lon, lat, elev, temp, slope, exposition, valley)*1
        terr_sky_view(slope, valley)
  rad_lw_bal(temp, rh, slope, valley, surface_type, surface_temp)*1
    rad_lw_in(temp, rh, slope, valley)*1.0
      rad_emissivity_air(temp, rh)*1
        pres_vapor_p(temp, rh)*1
          pres_sat_vapor_p(temp)*1.0
    rad_lw_out(surface_type, surface_temp)*1.0
```

# prepare 1 day of data
```{r}
# Metadata: inst/302_caldernklimawiese_complete_meta.xml
# Packaged five-minute teaching day extracted from the Caldern dataset.
input_day <- read.csv(caldern_path, na.strings = c("NA", "NULL"))
input_day <- input_day[!is.na(input_day$datetime) &
                         input_day$datetime >= "2017-06-30 00:00:00" &
                         input_day$datetime <= "2017-06-30 23:55:00", ]
```

# check .weather_station method with 3 day values and 3 night values
```{r, warning = FALSE}
# use selected complete day
input <- input_day

# Interpret timestamps as local station time for Caldern, Germany.
input$datetime <- as.POSIXlt(input$datetime, format = "%Y-%m-%d %H:%M:%S", tz = "Europe/Berlin")

# 3 day values and 3 night values
input <- input[c(1:3, 133:135), ]

# build_weather_station
weather_station <- build_weather_station(
  datetime = input$datetime,
  lon = 8.6832,
  lat = 50.8405,
  elev = 261,
  temp = input$Ta_2m,
  slope = 0,
  exposition = 0,
  valley = FALSE,
  surface_type = "field",
  surface_temp = input$Ta_2m,
  rh = input$Huma_2m,
  texture = "peat",
  moisture = input$water_vol_soil,
  soil_temp1 = input$Ts,
  soil_temp2 = input$Ta_2m,
  soil_depth1 = 0.25,
  soil_depth2 = 0
)

# show structure
str(weather_station)

# calculate
## soil
soil_heat_flux(weather_station)
  soil_thermal_cond(weather_station)

## rad
rad_bal(weather_station)
  rad_sw_bal(weather_station)
    rad_sw_out(weather_station)
      rad_sw_in(weather_station)
        rad_sw_toa(weather_station)
          sol_eccentricity(weather_station)
            sol_day_angle(weather_station)
              sol_julian_day(weather_station)
          sol_elevation(weather_station)
            sol_declination(weather_station)
              sol_ecliptic_length(weather_station)
                sol_medium_anomaly(weather_station)
            sol_hour_angle(weather_station)
              sol_medium_suntime(weather_station)
              sol_time_formula(weather_station)
        trans_gas(weather_station)
          trans_air_mass_abs(weather_station)
            trans_air_mass_rel(weather_station)
            pres_p(weather_station)
        trans_ozone(weather_station)
        trans_rayleigh(weather_station)
        trans_vapor(weather_station)
          hum_precipitable_water(weather_station)
        trans_aerosol(weather_station)
        terr_terrain_angle(weather_station)
          sol_azimuth(weather_station)
    rad_diffuse_out(weather_station)
      rad_diffuse_in(weather_station)
        terr_sky_view(weather_station)
  rad_lw_bal(weather_station)
    rad_lw_in(weather_station)
      rad_emissivity_air(weather_station)
        pres_vapor_p(weather_station)
          pres_sat_vapor_p(weather_station)
    rad_lw_out(weather_station)
```

# check .weather_station method with data for 1 day
```{r, warning = FALSE}
# use selected complete day
input <- input_day

# show structure
str(input)

# Interpret timestamps as local station time for Caldern, Germany.
input$datetime <- as.POSIXlt(input$datetime, format = "%Y-%m-%d %H:%M:%S", tz = "Europe/Berlin")

# build_weather_station
weather_station <- build_weather_station(
  datetime = input$datetime,
  lon = 8.6832,
  lat = 50.8405,
  elev = 261,
  temp = input$Ta_2m,
  slope = 0,
  exposition = 0,
  valley = FALSE,
  surface_type = "field",
  surface_temp = input$Ta_2m,
  rh = input$Huma_2m,
  texture = "peat",
  moisture = input$water_vol_soil,
  soil_temp1 = input$Ts,
  soil_temp2 = input$Ta_2m,
  soil_depth1 = 0.25,
  soil_depth2 = 0
)
```

## A function for plotting
```{r}
plotme <- function(topic, calculated, measured, legend_position) {
  values_combined <- c(calculated, measured)
  plot(
    calculated,
    xlab = "Day of year", ylab = "W/m2", main = topic,
    ylim = c(min(values_combined), max(values_combined))
  )
  lines(measured)
  legend(
    legend_position,
    c("calculated", "measured"),
    pch = c(1, NA),
    lty = c(0, 1)
  )
}
```

## Shortwave incoming
```{r}
topic <- "Shortwave incoming"
calculated <- rad_sw_in(weather_station) + rad_diffuse_in(weather_station)
measured <- input$rad_sw_in
legend_position <- "topright"
plotme(topic, calculated, measured, legend_position)
```

## Shortwave outgoing
```{r}
topic <- "Shortwave outgoing"
calculated <- rad_sw_out(weather_station) + rad_diffuse_out(weather_station)
measured <- input$rad_sw_out
legend_position <- "topright"
plotme(topic, calculated, measured, legend_position)
```

## Shortwave balance
```{r}
topic <- "Shortwave balance"
calculated <- rad_sw_bal(weather_station)
measured <- input$RsNet
legend_position <- "topright"
plotme(topic, calculated, measured, legend_position)
```

## Longwave incoming
```{r}
topic <- "Longwave incoming"
calculated <- rad_lw_in(weather_station)
measured <- input$LUpCo
legend_position <- "left"
plotme(topic, calculated, measured, legend_position)
```

## Longwave outgoing
```{r}
topic <- "Longwave outgoing"
calculated <- rad_lw_out(weather_station)
measured <- input$LDnCo
legend_position <- "left"
plotme(topic, calculated, measured, legend_position)
```

## Longwave balance
```{r}
topic <- "Longwave balance"
calculated <- rad_lw_bal(weather_station)
measured <- input$RlNet
legend_position <- "left"
plotme(topic, calculated, measured, legend_position)
```

## Radiation balance
```{r}
topic <- "Radiation balance"
calculated <- rad_bal(weather_station)
measured <- input$rad_net
legend_position <- "topright"
plotme(topic, calculated, measured, legend_position)
```

## Soil heat flux
```{r}
topic <- "Soil heat flux"
calculated <- soil_heat_flux(weather_station)
measured <- input$heatflux_soil
legend_position <- "topright"
plotme(topic, calculated, measured, legend_position)
```

# Examples
## .default method with optional arguments
```{r}
pres_p(0, 20)
pres_p(0, 20, p0 = 1013)
```

## .weather_station method with optional arguments
```{r}
weather_station <- build_weather_station(
  elev = 0,
  temp = 20
)
pres_p(weather_station)
pres_p(weather_station, p0 = 1013)
```
