---
title: "FieldClim Package Update Documentation"
author: "Konstantin Engelmayer"
date: "2024-08-10"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{FieldClim Package Update Documentation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
```


# Introduction

This document provides detailed documentation of the recent updates and
improvements made to various functions within the `fieldClim` package.
These updates primarily address issues related to radiation, sensible
heat flux, and latent heat flux functions. Additionally, the
standardization of energy flow signs (+/-) and enhanced documentation
for each function are also included.

### Overview of Changes:

-   R Studio help pages: Comprehensive documentation has been created
    for all functions in the package.
-   `plot_weather_station()`: New function added which plots all time
    series data of a `weather_station` object.
-   `bound_thermal_avg()`: Updated parameters for `turb_ustar(v1, z1)`
    to `v` and `z` to match required input.
-   `latent_priestley_taylor()`: Fixed an error in the formula that
    caused negative values where positive values were expected.
-   `latent_penman()`: Completely revised the formula to the correct
    version, and included soil flux for calculation.
-   `latent_monin()`: Added the possibility to set a cap for the
    stability parameter `s1` (measurement height / Monin-Obukhov length)
    to handle cases where the Monin-Obukhov length approaches zero,
    which results in really high `s1` values..
-   `latent_bowen()`: Added the possibility to set a cap parameter to
    prevent unrealistically high latent heat flux when (1 + B)
    approaches zero.
-   `sensible_bowen()`: Applied the same cap parameter as in
    `latent_bowen()` to avoid unrealistic values.
-   `sensible_monin()`: Applied the same cap as in `latent_monin()` to
    avoid unrealistic values.
-   `sensible_priestley_taylor()`: Corrected the sign direction for the
    calculation.
-   `solar_azimuth()`, `sol_hour_angle()`: Changed the timezone used
    from UTC to local time zone
-   `soil_heat_flux()`: Reversed the sign for the flux direction to
    ensure consistency.

## 1. Added proper Help Page Details

The biggest part of this update includes the improved R Studio Help
pages. The help documentation for each function in RStudio now features
a comprehensive "Details" section that provides an in-depth explanation
of the function's purpose and operation. For functions based on
mathematical formulas, these formulas are now included and thoroughly
explained within this section. This enhancement significantly improves
user-friendliness, making the package accessible to users who were not
involved in its development. Additionally, each help page now includes
example code and specifies the unit of the output value, further aiding
in the practical application of the functions.

<!-- Figure removed from self-contained HTML: Example R Studio help page of the sensible_monin() -->

**Figure 1:** Example R Studio help page of the `sensible_monin()`
function

## 2. `plot_weather_station()`

The `plot_weather_station()` function is a new addition to the package
that provides a simple and intuitive way to visualize time series data
stored within a `weather_station` object. The function can either plot a
single specified variable or generate a grid of plots for all
time-series variables within the object.
- Functionality: - **Single
Variable Plot**: Users can specify a single variable by name to plot its
time series.
- **Multiple Variable Plot**: If no variable is specified,
the function automatically detects all time-series variables (those with
the same length as the datetime variable) and generates a grid of
plots.
- **Automatic Labeling**: The function includes a built-in
mapping of variable names to descriptive labels with appropriate units,
ensuring that plots are informative and easy to interpret.
- Usage:

This will create a plot of the air temperature time series.

```{r, eval=FALSE}
plot_weather_station(weather_station, variable_name = "temp")
```

<!-- Figure removed from self-contained HTML.
-->

This will create a grid of plots for all available time-series variables
within the weather_station object.

```{r, eval=FALSE}
plot_weather_station(weather_station)
```

<!-- Figure removed from self-contained HTML.
-->

## 3. `bound_thermal_avg()`

The function is calling the `turb_ustar()` function. However the
arguments of the function are not called `v1` and `z1` but `v` and `z`.
This is now fixed in the new Version.

```{r, eval=FALSE}
bound_thermal_avg <- function(v, z, temp_change_dist, t_pot_upwind, t_pot, lapse_rate,
                              surface_type = NULL, obs_height = NULL) {
  # Calculate ustar
  if (!is.null(obs_height)) {
    ustar <- turb_ustar(v1 = v, z1 = z, obs_height = obs_height) # v1 -> v, z1 -> z 
  } else if (!is.null(surface_type)) {
    ustar <- turb_ustar(v1 = v, z1 = z, surface_type = surface_type) # v1 -> v, z1 -> z 
  } else {
    print("The input is not valid. Either obs_height or surface_type has to be defined.")
  }
  (ustar / v) * ((temp_change_dist * abs(t_pot_upwind - t_pot)) / abs(lapse_rate))**0.5
}
```

## 4. `latent_priestley_taylor()`

Here an error was fixed in the formula that caused negative values where
positive values were expected.

```{r, eval=FALSE}
# Old formula
out <- alpha_pt * sc * (-rad_bal - soil_flux) / (sc + gam)

# New formula
out <- alpha_pt * (delta / (delta + gamma)) * (rad_bal - soil_flux)
```

## 5. `latent_penman()`

For this function the formula was revised completely. Before the `water`
package was used which is now deprecated. The function now includes soil
flux, surface resistance and aerodynamic resistance for calculation.
Values for surface resistance where taken from literature, same as the
formula and other calculations needed: \*- Monteith, John L., Mike H.
Unsworth, and Ann Webb. "Principles of environmental physics." Quarterly
Journal of the Royal Meteorological Society 120.520 (1994): 1699. The
parameters `obs_height` and `surface_type` to calculate the aerodynamic
resistance and get the right value for surface resistance. The formula
used looks like this:

[\\\[ Q_e = \\frac{\\Delta \\times (R_n - G) + \\gamma \\times
\\left(\\frac{c_p \\times \\rho}{r_a}\\right) \\times (e_s -
e_a)}{\\Delta + \\gamma \\times \\left(1 + \\frac{r_s}{r_a}\\right)}
\\\]]{.math .display}

where:

- \(Q_e\) is the latent heat flux (W/m²).
-
\(\\Delta\) is the slope of the vapor pressure curve
(kPa/°C).
- \(R_n\) is the net radiation (W/m²).
-
\(G\) is the soil heat flux density (W/m²).
-
\(\\gamma\) is the psychrometric constant (kPa/°C).
-
\(c_p\) is the specific heat of air at constant
pressure (J/kg/°C).
- \(\\rho\) is the air density
(kg/m³).
- \(r_a\) is the aerodynamic resistance
(s/m).
- \(r_s\) is the surface resistance (s/m).
-
\(e_s\) is the saturation vapor pressure (kPa).
-
\(e_a\) is the actual vapor pressure (kPa). \]

<!-- Figure removed from self-contained HTML: Comparison of results of old and new latent_penman() function using -->

**Figure 2:** Comparison of results of old and new `latent_penman()`
function using 'energie_bil_wiese.csv'

## 6. `latent_monin()`

The function `latent_monin()` includes the stability correction for
humidity. The calculation of this correction is based on the stability
parameter \(s_1\), which is defined as:

[\\\[ s_1 = \\frac{\\text{measurement height}}{\\text{Monin-Obukhov
length}} \\\]]{.math .display}

However when the Monin-Obukhov length approaches zero, [\\(s_1\\)]{.math
.inline} can become extremely large, leading to unrealistic values in
further calculations. To address this, a possibility to set a cap for
\(s_1\) was added. The cap is `NULL` by default and
can be set with `cap = x`.

In the function the cap use looks like this:

```{r, eval=FALSE}
if(!is.null(cap)){
    # Apply cap
    s1 <- pmax(pmin(s1, cap), -cap)
  }
```

Also an error was fixed in the formula that caused negative values where
positive values were expected.

<!-- Figure removed from self-contained HTML: Comparison of results of old and new latent_monin() function using cap = 20 and -->

**Figure 3:** Comparison of results of old and new `latent_monin()`
function using `cap = 20` and 'energie_bil_wiese.csv'

## 7. `latent_bowen()`

The latent heat flux \(Q_e\) in the Bowen ratio
method is calculated using the formula:

[\\\[ Q_e = \\frac{Q_n - Q_g}{1 + B} \\\]]{.math .display}

Where:

- \(Q_e\) is the latent heat flux (W/m²).
-
\(Q_n\) is the net radiation (W/m²).
-
\(Q_g\) is the soil heat flux (W/m²).
-
\(B\) is the Bowen ratio, defined as the ratio of
sensible heat flux to latent heat flux. Here there was a similar problem
as for the `latent_monin()` function. When [\\((1 + B)\\)]{.math
.inline} approaches zero, \(Q_e\) can become
extremely large, leading to unrealistic values. To prevent this, a cap
\(C\) is introduced ensuring that [\\((1 +
B)\\)]{.math .inline} does not approach zero by maintaining a minimum
value set by the cap. The cap is `NULL` by default and can be set with
`cap = x`.

In the function the cap use looks like this:

```{r, eval=FALSE}
denominator <- 1 + bowen_ratio
  if(!is.null(cap)){
    denominator[abs(denominator) < cap] <- cap * sign(denominator[abs(denominator) < cap])
  }
```

Also an error was fixed in the formula that caused negative values where
positive values were expected.

<!-- Figure removed from self-contained HTML: Comparison of results of old and new latent_bowen() function using cap = 1 and -->

**Figure 4:** Comparison of results of old and new `latent_bowen()`
function using cap = 1 and 'energie_bil_wiese.csv'

## 8. `sensible_monin()` `sensible_bowen()`

For these functions the same new `cap` parameter was added to the
function as for `latent_monin()` and `latent_bowen()`.

## 9. `solar_azimuth()` `sol_hour_angle()`

In the previous implementation of `solar_azimuth()` and
`sol_hour_angle()`, the `solar_medium_suntime()` function, which
converts time to UTC, was used. However, these calculations require
local time. Therefore, the use of `solar_medium_suntime()` was removed,
and local time is now directly utilized. This error caused time shifts
in all radiation calculations, but with these changes this issue has now
been resolved.

## 10. `soil_heat_flux()`

Here the sign for the flux direction was reversed to ensure consistency.

```{r, eval=FALSE}
-thermal_cond * (soil_temp1 - soil_temp2) / (soil_depth1 - soil_depth2)
```
