---
title: "Scientific background for fieldClim heat-flux methods"
author: "Jörg Bendix, Chris Reudenbach"
date: "2026-05-28"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Scientific background for fieldClim heat-flux methods}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Purpose of this vignette

This vignette provides the scientific background for the heat-flux methods implemented in `fieldClim`. It is not a numerical benchmark and it is not a replacement for method-specific validation with independent flux measurements. Its purpose is narrower and more important for package use: it explains why several approximate micrometeorological methods are implemented side by side, what each method estimates, what data it needs, and which assumptions must be visible when the results are interpreted.

`fieldClim` is designed for stations that measure the key near-surface energy-balance drivers but do not include full eddy-covariance instrumentation. In such settings, radiation, soil heat flux, air temperature, humidity and wind profiles can still be used to estimate or diagnose turbulent heat-flux pathways. The package therefore focuses on physically interpretable approximation methods for station-based microclimate analysis, not on high-frequency eddy-covariance processing.

Eddy covariance remains the direct turbulent-flux reference method in many comparison studies. The methods implemented in `fieldClim` should therefore be read as model-based or diagnostic estimates from standard station data. They are not interchangeable measurements of the same object. Some methods close the available energy by construction, some provide only latent heat flux, and some are diagnostic profile or gradient methods that are not expected to close the energy balance.

# Surface energy balance and package sign convention

The common reference point is the surface energy balance. Microclimatological and micrometeorological texts use partly different symbols for the same physical quantities. `fieldClim` uses technical field and output names such as `rad_bal`, `soil_flux`, `sensible_*` and `latent_*`. These are implementation names, not a separate theoretical notation.

```{r sign-convention-table, echo = FALSE}
sign_convention <- data.frame(
  Quantity = c(
    "Net radiation",
    "Soil heat flux",
    "Sensible heat flux",
    "Latent heat flux",
    "Available turbulent energy"
  ),
  `Common notation` = c(
    "Q* or Rn",
    "B or G",
    "H or L",
    "LE or V",
    "Q* - B or Rn - G"
  ),
  `fieldClim field / output` = c(
    "rad_bal",
    "soil_flux",
    "sensible_*",
    "latent_*",
    "rad_bal - soil_flux"
  ),
  `Positive direction / meaning in fieldClim` = c(
    "net radiative energy input at the surface",
    "heat flux into the soil",
    "heat flux away from the surface",
    "heat flux away from the surface",
    "energy available for sensible and latent heat fluxes"
  ),
  check.names = FALSE
)

knitr::kable(sign_convention)
```

With storage omitted, the surface energy balance is written here as:

$$
R_n = G + H + LE
$$

Therefore the energy available for turbulent heat fluxes is:

$$
R_n - G = H + LE
$$

In the notation used in the accompanying theory material, the same balance is:

$$
Q^* - B = L + V
$$

A residual latent heat flux follows directly from the same balance:

$$
LE = R_n - G - H
$$

This convention is used throughout the package: `soil_flux` is consumed as a positive heat flux into the soil, while positive sensible and latent heat fluxes are interpreted as fluxes away from the surface.

# Direct flux measurement as reference, not as package scope

Eddy covariance (EC) uses high-frequency covariance between vertical wind and scalar fluctuations to estimate turbulent fluxes. It is often treated as the most direct field reference for sensible and latent heat fluxes because it measures turbulent exchange rather than inferring it from a mean-gradient or residual relation. However, EC is not a trivial truth source. Energy-balance non-closure is a persistent issue in EC datasets, and closure depends on footprint, heterogeneity, averaging period, storage terms, advection and instrumentation.

For `fieldClim`, this means two things. First, the package does not implement EC processing. A field called `EC` in station metadata must not be interpreted as eddy covariance unless the data actually contain high-frequency sonic and gas-analyser measurements. Second, EC comparison literature is still essential background because it tells us how approximate station-based methods behave when compared to direct turbulent-flux systems.

# Method families implemented in fieldClim

```{r method-family-table, echo = FALSE}
methods <- data.frame(
  Method_family = c(
    "Priestley-Taylor",
    "Bulk-Residual",
    "Bowen ratio",
    "Penman / Penman-type",
    "Monin-Obukhov / profile",
    "Radiation and soil helpers"
  ),
  Package_functions = c(
    "latent_priestley_taylor(), sensible_priestley_taylor()",
    "sensible_bulk(), latent_bulk_residual(), turb_flux_bulk_residual()",
    "sensible_bowen(), latent_bowen()",
    "latent_penman()",
    "sensible_monin(), latent_monin(), turb_flux_stability()",
    "rad_*(), sol_*(), trans_*(), soil_*()"
  ),
  Main_output = c(
    "H and LE partition of Rn - G",
    "H from bulk resistance; LE as residual",
    "H and LE from gradient ratio",
    "LE only",
    "diagnostic H and LE from profile/stability logic",
    "radiation, soil heat flux, helper quantities"
  ),
  Closure_status = c(
    "closes Rn - G by construction",
    "closes Rn - G by construction",
    "closes Rn - G only for finite uncapped denominators",
    "LE only; no paired H",
    "diagnostic only; not forced to close Rn - G",
    "provides inputs and consistency checks"
  ),
  Key_assumption = c(
    "empirical Priestley-Taylor coefficient and available energy",
    "neutral wind-dependent resistance and residual closure",
    "temperature and humidity gradients represent turbulent partitioning",
    "energy term plus aerodynamic vapour-pressure term",
    "profile gradients and stability classification are meaningful",
    "radiation and soil parameters are internally consistent"
  )
)

knitr::kable(methods)
```

The methods therefore represent approximation families. The correct question is not which method is the real one. The correct question is which assumptions are acceptable for the available station data and which diagnostic warnings appear in specific time steps.

# Priestley-Taylor path

The Priestley-Taylor path estimates latent heat flux from available energy and a coefficient. The implemented teaching path follows the structure:

$$
LE_{PT} =
\alpha
\frac{\Delta}{\Delta + \gamma}
(R_n - G)
$$

The corresponding sensible heat flux is calculated as the remaining available energy:

$$
H_{PT} = (R_n - G) - LE_{PT}
$$

Therefore the closure invariant is exact:

$$
H_{PT} + LE_{PT} = R_n - G
$$

This is why `turb_flux_calc(weather_station, pt_only = TRUE)` is useful as a beginner-safe workflow. It needs fewer profile assumptions than Bowen or Monin-Obukhov and avoids optional Penman, Bowen, Bulk and Monin paths.

The limitation is also clear. Priestley-Taylor hides much of the surface-atmosphere coupling in the coefficient `alpha` and in helper terms such as the saturation-curve slope and psychrometric constant. It is stable and transparent for teaching, but it is not a direct measurement of turbulent exchange.

# Penman-type latent heat path

The Penman family combines an energy term and an aerodynamic term. In `fieldClim`, `latent_penman()` is a latent-heat method only. It does not return a paired sensible heat flux.

The package convention keeps the energy term as:

$$
R_n - G
$$

The implemented Penman path uses saturation vapour pressure and actual vapour pressure helpers that return hPa, converts them internally to kPa for the aerodynamic vapour-pressure deficit, and keeps `Delta` and `gamma` on the same kPa scale:

$$
VPD_{kPa} = \frac{e_s - e_a}{10}
$$

The correction is important because mixing hPa vapour pressure deficit with kPa-scale `Delta` and `gamma` inflates the aerodynamic term. After the unit correction, the method remains LE-only, field-surface mapping is supported, and the output should still be interpreted as a modelled latent heat flux, not as an energy-closing pair.

The relevant package distinction is:

$$
\text{Penman path} = LE \text{ only}
$$

There is no package-defined `H_Penman`.

# Bowen-ratio path

The Bowen-ratio method partitions available energy using the ratio of sensible to latent heat flux:

$$
\beta = \frac{H}{LE}
$$

If the ratio is known, the energy balance gives:

$$
H =
\frac{\beta}{1 + \beta}
(R_n - G)
$$

and:

$$
LE =
\frac{1}{1 + \beta}
(R_n - G)
$$

For finite uncapped denominators, the closure invariant is exact:

$$
H + LE = R_n - G
$$

The implemented `fieldClim` Bowen ratio is based on a potential-temperature gradient and an absolute-humidity gradient:

$$
\beta =
\gamma_{code}
\frac{\Delta \theta / \Delta z}{\Delta AH / \Delta z}
$$

where:

$$
\gamma_{code} =
0.00066
(1 + 0.000946 t_1)
$$

The important documentation point is that this is not described as a generic symbol-only formula. The implementation converts `t1` and `t2` to potential temperature, converts `hum1` and `hum2` from relative humidity to absolute humidity, and then forms the gradient ratio.

The Bowen path is useful because it directly links two measured profile gradients to energy partitioning. It is also fragile because small humidity gradients, sign changes and near-zero values of `1 + beta` can create very large fluxes. The `cap` parameter is a numerical safeguard for near-zero denominators. When the cap is active, exact closure is no longer guaranteed.

Therefore:

$$
H_{Bowen} + LE_{Bowen} = R_n - G
$$

only for finite uncapped denominators.

# Bulk-Residual path

The Bulk-Residual path combines a wind- and temperature-gradient estimate of sensible heat flux with a residual latent heat flux. It is not a direct flux measurement. It is a station-data approximation in the bulk aerodynamic / aerodynamic-resistance family.

The sensible heat calculation follows the general aerodynamic resistance form:

$$
H_{bulk} =
\rho c_p
\frac{\Delta T}{r_a}
$$

In the current package implementation:

$$
\Delta T = T_1 - T_2
$$

and:

$$
r_a =
\frac{\ln(z_2 / z_1)}
{k \bar{u}}
$$

Here `T1` and `T2` are lower and upper air temperatures, `z1` and `z2` are the corresponding measurement heights, `k` is the von Karman constant and `u_mean` is the observed wind-speed scale used by the package. If two wind heights are available, `u_mean` is the arithmetic mean of `v1` and `v2`; otherwise `v1` is used.

This formulation should be documented neither as a freely invented proxy nor as a full Monin-Obukhov or SEBAL/METRIC method. The established method family uses aerodynamic resistance and wind-driven turbulent exchange. The package implementation is a simplified neutral station-data approximation because the teaching dataset provides wind speed but not directly measured friction velocity.

A more complete surface-layer formulation would often use friction velocity, roughness length, displacement height and stability corrections. Those quantities are not required by `sensible_bulk()`.

The residual latent heat flux is:

$$
LE_{res} =
R_n - G - H_{bulk}
$$

Therefore the Bulk-Residual workflow closes available energy by construction:

$$
H_{bulk} + LE_{res} = R_n - G
$$

This closure is algebraic. It does not prove that `H_bulk` is a perfect physical estimate. It means that the residual latent heat flux inherits all errors in `Rn`, `G` and the bulk sensible heat estimate.

# Wind speed, friction velocity and two-height station data

The use of wind is not an add-on. It is part of the aerodynamic-transfer logic. Stronger turbulent exchange lowers aerodynamic resistance, so the same temperature difference produces a larger sensible heat flux when wind-driven mixing is stronger.

The package uses observed wind speed because that is what standard station data provide. If only one wind speed is available, a friction-velocity based method cannot be used without roughness assumptions. If two wind heights are available, friction velocity can be estimated under a neutral log-profile assumption:

$$
u_* =
k
\frac{u_2 - u_1}
{\ln(z_2 / z_1)}
$$

and then a neutral heat-transfer resistance can be written as:

$$
r_{ah} =
\frac{\ln(z_2 / z_1)}
{k u_*}
$$

This is a possible extension, but it is not automatically more robust. If the wind-speed difference is small, noisy or changes sign, the estimated friction velocity becomes unstable. For a teaching and diagnostic station workflow, using observed mean wind speed is transparent and robust, provided the approximation is documented.

A stability-aware extension is also possible because two heights allow a gradient Richardson number to be estimated:

$$
Ri_g =
\frac{g}{\bar{\theta}}
\frac{\Delta \theta / \Delta z}
{(\Delta u / \Delta z)^2}
$$

With wind speed rather than wind vector components, this is a speed-shear approximation. It can be used as a guard to identify stable or weak-shear cases where a simple neutral bulk estimate should be treated cautiously or set to `NA`. Such a guard is not the same as a full Monin-Obukhov correction.

# Monin-Obukhov and profile methods

Monin-Obukhov similarity theory and related profile-gradient methods attempt to represent turbulent transfer through vertical gradients, roughness, stability and surface-layer scaling. In package interpretation, the Monin path is diagnostic-only.

The diagnostic rule is:

$$
H_{MO} + LE_{MO}
$$

is not required to equal:

$$
R_n - G
$$

This is not a defect. It is a consequence of method class. A Monin/profile method estimates turbulent fluxes from profile and stability assumptions, while PT, Bowen and Bulk-Residual are explicit energy-partition or residual workflows.

The practical issue is numerical robustness. Zero gradients, very small wind shear, invalid height relationships and unstable stability functions can produce extreme outputs. For this reason, Monin-Obukhov outputs should be interpreted together with stability classification and diagnostic warnings. They should not be silently normalized to available energy.

# Evidence from intercomparison literature

The implemented method families are not isolated package inventions. They belong to a large literature comparing direct turbulent-flux measurements with energy-balance, gradient, aerodynamic, residual and evaporation-model approaches.

A recent year-long grassland comparison by Billesbach et al. compared eddy covariance, energy-balance/Bowen-ratio, residual-energy and modified Bowen-ratio methods for sensible and latent heat flux over a mixed grass prairie. This is directly relevant to `fieldClim` because it treats the same family structure: direct EC, energy-balance closure, residual methods and Bowen-type gradient logic.

FAO-56 provides the standard reference for Penman-Monteith style evapotranspiration calculations from meteorological data. It explicitly distinguishes aerodynamic resistance and surface resistance, and it shows how aerodynamic resistance can be reduced to wind-dependent forms under standardized assumptions.

Priestley and Taylor introduced a large-scale energy-parameter approach for evaporation and surface heat fluxes, which motivates the PT family used in many simplified evaporation and energy-balance applications.

Penman combined energy and aerodynamic reasoning for natural evaporation from open water, bare soil and grass. The package Penman path follows that method family in a simplified latent-heat implementation.

Zhao et al. compared EC, Bowen-ratio energy balance and aerodynamic methods for paddy fields and emphasized the role of measurement-height differences, growth stage, weather and atmospheric stability. This is directly relevant to station workflows with two measurement heights.

Prueger and Kustas discuss aerodynamic methods for estimating turbulent fluxes in agricultural systems, including flux-gradient approaches. This provides method-family background for profile and aerodynamic transfer estimates.

Kim et al. validated sensible heat flux estimates from a bulk transfer method against a three-dimensional ultrasonic anemometer or scintillometer at multiple sites. This is especially relevant to the question of whether bulk-transfer estimates are empirically compared against sonic-anemometer-type flux measurements.

Ala-Konni et al. compared bulk transfer models with eddy-covariance turbulent heat fluxes over seasonal lake ice for several ice seasons. This demonstrates the broader use of bulk-transfer model validation against EC, even though the surface type differs from the Caldern meadow example.

Sumner and Jacobs compared Penman-Monteith, Priestley-Taylor, reference evapotranspiration and pan evaporation approaches for pasture evapotranspiration. This is relevant for interpreting PT and Penman-type methods as practical evapotranspiration approximations rather than direct turbulent-flux measurements.

Tomlinson compared Bowen-ratio, eddy-correlation and Priestley-Taylor approaches in an evapotranspiration context and explicitly connects energy-budget closure with sensible and latent heat flux estimation.

The conclusion is not that one station method is universally correct. The conclusion is that the implemented approaches belong to evaluated micrometeorological method families. They differ in data requirements, closure behaviour and sensitivity to gradients, wind, stability and surface assumptions.

# Practical interpretation in fieldClim

The methods should be read as complementary diagnostics.

Priestley-Taylor is the most stable teaching path. It partitions available energy with few profile assumptions and closes exactly.

Bulk-Residual is transparent and uses directly available station data: temperature gradient, wind speed, net radiation and soil heat flux. Its latent heat flux is residual and therefore closes exactly, but the sensible heat component is a simplified neutral bulk estimate.

Bowen is a gradient-ratio partition method. It is physically meaningful when gradients are well resolved, but it is highly sensitive to small humidity gradients and near-singular denominators.

Penman is a latent-heat-only combination method. It should be used as an LE estimate, not as an energy-closing H/LE pair.

Monin-Obukhov is a diagnostic profile/stability path. It is methodologically important, but its output should be checked for zero-gradient, low-wind and stability singularities. It should not be forced to close available energy.

# Recommended wording for package documentation

For method summaries, use symmetrical language. Do not over-defend one method while treating the others as exact.

A robust summary is:

> `fieldClim` implements common micrometeorological approximation families for heat-flux analysis from weather-station data. Priestley-Taylor, Bowen-ratio, Penman-type, bulk-resistance/residual and Monin-Obukhov/profile methods differ in data requirements and assumptions. Some methods close the available energy by construction; others are LE-only or diagnostic profile estimates. Eddy covariance and 3D sonic-anemometer studies provide empirical reference comparisons, but EC itself has known energy-balance closure limitations.

# References

Ala-Konni, J., Mammarella, I., Ojala, A., et al. (2022). Validation of turbulent heat transfer models against eddy-covariance flux measurements over a seasonally ice-covered lake. *Geoscientific Model Development*, 15, 4739-4755. https://doi.org/10.5194/gmd-15-4739-2022

Allen, R. G., Pereira, L. S., Raes, D., & Smith, M. (1998). *Crop evapotranspiration: Guidelines for computing crop water requirements*. FAO Irrigation and Drainage Paper 56. Food and Agriculture Organization of the United Nations.

Bastiaanssen, W. G. M., Menenti, M., Feddes, R. A., & Holtslag, A. A. M. (1998). A remote sensing surface energy balance algorithm for land (SEBAL). 1. Formulation. *Journal of Hydrology*, 212-213, 198-212.

Billesbach, D. P., Arkebauer, T. J., Walters, D. T., & Verma, S. B. (2024). Intercomparison of sensible and latent heat flux measurements from combined eddy covariance, energy balance, and Bowen ratio methods above a grassland prairie. *Scientific Reports*, 14, 21486. https://doi.org/10.1038/s41598-024-67911-z

Foken, T. (2008). The energy balance closure problem: An overview. *Ecological Applications*, 18(6), 1351-1367. https://doi.org/10.1890/06-0922.1

Kim, M. S., Kwon, B. H., & Kim, J. (2019). Estimation of sensible heat flux and atmospheric boundary layer height using an unmanned aerial vehicle. *Atmosphere*, 10(7), 363. https://doi.org/10.3390/atmos10070363

Penman, H. L. (1948). Natural evaporation from open water, bare soil and grass. *Proceedings of the Royal Society of London. Series A*, 193, 120-145. https://doi.org/10.1098/rspa.1948.0037

Priestley, C. H. B., & Taylor, R. J. (1972). On the assessment of surface heat flux and evaporation using large-scale parameters. *Monthly Weather Review*, 100(2), 81-92. https://doi.org/10.1175/1520-0493(1972)100<0081:OTAOSH>2.3.CO;2

Prueger, J. H., & Kustas, W. P. (2005). Aerodynamic methods for estimating turbulent fluxes. In J. L. Hatfield & J. M. Baker (Eds.), *Micrometeorology in agricultural systems*. Agronomy Monograph 47. ASA, CSSA and SSSA.

Shi, T., Guan, D., Wu, J., Wang, A., Jin, C., & Han, S. (2008). Comparison of methods for estimating evapotranspiration rate of dry forest canopy: Eddy covariance, Bowen ratio energy balance, and Penman-Monteith equation. *Journal of Geophysical Research: Atmospheres*, 113, D19116. https://doi.org/10.1029/2008JD010174

Sumner, D. M., & Jacobs, J. M. (2005). Utility of Penman-Monteith, Priestley-Taylor, reference evapotranspiration, and pan evaporation methods to estimate pasture evapotranspiration. *Journal of Hydrology*, 308(1-4), 81-104. https://doi.org/10.1016/j.jhydrol.2004.10.023

Tomlinson, S. A. (1996). *Comparison of Bowen-ratio, eddy-correlation, and weighing-lysimeter evapotranspiration measurements in a semiarid rangeland*. U.S. Geological Survey Water-Resources Investigations Report 96-4081.

Zhao, Y., et al. (2023). Comparing the eddy covariance and gradient methods for estimating evapotranspiration in paddy fields. *Agricultural Water Management*, 284, 108332. https://doi.org/10.1016/j.agwat.2023.108332
